---
title: "Transform"
---

## Overview

The goal of this script is to transform the curated State of the Union data into a format that is suitable for analysis. For reference, the research question is:

> Are there any notable differences in the language used by presidents from pre- and post-9/11? Starting after WWII and up to the present?

This is exploratory research and the analysis will be conducted using a variety of methods including:

- Sentiment analysis *(demo here)*
- Topic modeling
- Word co-occurrence

The aim is to gain possible insights into changes in presidential rhetoric before the historical event of 9/11 and after.

To facilitate the transformation process, I will load various R packages.

```{r load-packages}
#| label: load-packages
#| message: false

# Load packages
library(readr)        # reading/ writing datasets
library(knitr)        # for `kable()` function
library(dplyr)        # basic data manipulation
library(lubridate)    # date manipulation
library(tidytext)     # text manipulation
library(textdata)     # sentiment lexicons
library(qtalrkit)     # data dictionary creation
library(ggplot2)      # plotting
library(fs)           # file system operations
```

## Description

The data is a collection of State of the Union addresses from 1945 to present. The data dictionary for this dataset is seen in @tbl-sotu-curated-dd.

```{r}
#| label: tbl-sotu-curated-dd
#| tbl-cap: Data dictionary for the State of the Union dataset
#| echo: false
#| message: false

# Load the data dictionary
read_csv("../data/derived/sotu_curated_dd.csv") |>
  kable()
```

The data dictionary shows that we have a number of variables that we can use for analysis. The `text` variable will be the most important for the analysis. It contains the actual text of the State of the Union addresses. The `date` variable will be useful for filtering the data for only addresses after WWII (1945-). The other variables may be useful for additional follow up analysis.

As it stands, the curated dataset has the `text` as the unit of observation. This corresponds to the individual addresses.

## Structure

The first step in the transformation process is to load the data and inspect the structure of the dataset.

```{r}
#| label: load-data
#| message: false

# Load the data
sotu_tbl <- read_csv("../data/derived/sotu_curated.csv")

# Preview
glimpse(sotu_tbl)
```

We now see the variables (as described in the data dictionary) as part of the `sotu_tbl` data frame. There are 241 rows and 7 columns.

The next step is to transform the data into a format that is suitable for analysis. Since there are a number of different analyses that will be conducted, I will create a number of different datasets that are suitable for each analysis.

**Sentiment analysis**

The first analysis will be sentiment analysis. This will involve using the `bing` lexicon to calculate the sentiment of each word in the `text` variable. The `bing` lexicon is a simple lexicon that classifies words as either positive or negative. The `tidytext` package will be used to tokenize the words in the `text` variable. Then the `bing` lexicon will be joined to the tokenized words to calculate the sentiment of each word.

An idealized structure for the sentiment analysis dataset is seen in @tbl-sentiment-ideal.

| variable | name | type | description |
|----------|------|-------------|--|
| address_id | Address ID | integer | Unique identifier for each address |
| address_year | Address Year | integer | Year of the address (e.g. 1945, 1978, etc.) |
| period | Period | character | Pre- or post-9/11 |
| word | Word tokens | character | Individual words from the `text` variable |
| sentiment | Sentiment class | character | Sentiment of the word (positive or negative) |

: Idealized structure for the sentiment analysis dataset {#tbl-sentiment-ideal tbl-colwidths="[15, 15, 15, 55]"}


**Topic modeling**

For topic modeling, the `text` variable will be tokenized into words. In the analysis, we will likely want the frequency of words and the TF-IDF of words (to weight the importance of words). It will be important to have the `address_id` and `address_year` variables to keep track of the address and year of the address as we will want to split the dataset by `period` before the topic modeling analysis.

An idealized structure for the topic modeling dataset is seen in @tbl-topic-ideal.

| variable | name | type | description |
|----------|------|-------------|--|
| address_id | Address ID | integer | Unique identifier for each address |
| address_year | Address Year | integer | Year of the address (e.g. 1945, 1978, etc.) |
| period | Period | character | Pre- or post-9/11 |
| type | Word type | character | Individual words from the `text` variable |
| frequency | Frequency | integer | Frequency of the word type in the `text` variable |
| tf_idf | TF-IDF | numeric | TF-IDF of the word type in the `text` variable |

: Idealized structure for the topic modeling dataset {#tbl-topic-ideal tbl-colwidths="[15, 15, 15, 55]"}

**Word Co-occurrence**

For word co-occurrence, we will want to calculate the co-occurrence of words in the `text` variable. This will involve tokenizing the `text` variable into words and then calculating the co-occurrence of words. The `address_id` and `address_year` variables will be important to keep track of the address and year of the address as we will want to split the dataset by `period` before the word co-occurrence analysis.

An idealized structure for the word co-occurrence dataset is seen in @tbl-co-occurrence-ideal.

| variable | name | type | description |
|----------|------|-------------|--|
| address_id | Address ID | integer | Unique identifier for each address |
| address_year | Address Year | integer | Year of the address (e.g. 1945, 1978, etc.) |
| period | Period | character | Pre- or post-9/11 |
| word_ref | Reference word | character | Reference word from the `text` variable |
| word_co | Co-occurring word | character | Co-occurring word from the `text` variable |
| co_occurrence | Co-occurrence | integer | Co-occurrence of the reference and co-occurring word (as PMI score) |

: Idealized structure for the word co-occurrence dataset {#tbl-co-occurrence-ideal tbl-colwidths="[15, 15, 15, 55]"}

## Transform

Let's aim at transforming the data into the idealized structures for each analysis. We will start with the sentiment analysis dataset.

### Sentiment Analysis

<!--
| variable | name | type | description |
|----------|------|-------------|--|
| address_id | Address ID | integer | Unique identifier for each address |
| address_year | Address Year | integer | Year of the address (e.g. 1945, 1978, etc.) |
| period | Period | character | Pre- or post-9/11 |
| word | Word tokens | character | Individual words from the `text` variable |
| sentiment | Sentiment class | character | Sentiment of the word (positive or negative) |
-->

Let's start by setting up our metadata. I will do this so that we can easily keep track of the variables in the dataset once we move to tokenize the text.

The metadata columns to create are:

- `address_id`: Unique identifier for each address
- `address_year`: Year of the address
- `period`: Pre- or post-9/11


First, I will extract the year from the `date` variable and add it to a column `address_year`. The `date` variable has been read in (with `readr`) as a 'date' class. This means that it is possible to extract the year from the date using the `year()` function from the `lubridate` package.

```{r}
#| label: sentiment-metadata-address-year

# Add the address_year
sotu_tbl <-
  sotu_tbl |>
  mutate(address_year = year(date))

# Preview
glimpse(sotu_tbl)
```

The next step is to work with the `year` column. We are going to first filter the rows such that we only have addresses after WWII (1945-).

```{r}
#| label: sentiment-metadata-filter-years

# Filter the data for addresses after WWII
sotu_tbl <-
  sotu_tbl |>
  filter(address_year >= 1945)

# Preview
glimpse(sotu_tbl)
```

We can now see we've dropped from 241 addresses to 86.

Then we will create the `period` column using the `year` column to add `pre_9_11` or `post_9_11` to the `period` column with the `case_when()` function. The cutoff year for pre-9/11 is 2001.

```{r}
#| label: sentiment-metadata-period

# Add the period
sotu_tbl <-
  sotu_tbl |>
  mutate(period = case_when(
    address_year < 2001 ~ "pre",
    address_year >= 2001 ~ "post"
  ))

# Preview
glimpse(sotu_tbl)
```

Finally, we will add the `address_id`, all we need to do is to create a sequence of numbers from 1 to the number of rows in the dataset. A shortcut for this operation is the `row_numbers()` function from the `dplyr` package.

```{r}
#| label: sentiment-metadata-address-id

# Add the address_id
sotu_tbl <-
  sotu_tbl |>
  mutate(address_id = row_number())

# Preview
glimpse(sotu_tbl)
```

Now, we are in a position to tokenize the `text` variable. But before we do that, we need to decided if we are keeping the other variables or dropping them. Some of the variables may be useful for additional analysis, either to follow up on potential patterns or to use to filter the dataset even more.

For example, say see a trend in more negative sentiment in the post-9/11 addresses. We may want to investigate if this is related to the president's party or even the president. In this case, we would want to keep the `president` and `party` variables.

On the other hand, a variable like `delivery` may turn out to be a variable we don't want to have add additional complexity to the analysis. In this case, we might want to decide which value (level) of delivery to keep. Let's take a quick look at a frequency table of the `delivery` variable.

```{r}
#| label: tbl-delivery-freq
#| tbl-cap: Frequency table of the `delivery` variable

# Frequency table of the `delivery` variable
sotu_tbl |>
  count(delivery) |>
  kable()
```

In @tbl-delivery-freq, we see that the great majority of the addresses in the dataset are 'spoken'. Whatever potential differences from the 'written' modality may not be worth the additional complexity in the analysis.

With this in mind, we will filter the dataset to only keep `delivery == 'spoken'` and then drop:

- `delivery`
- `date`
- `address_type`
- `first_name`

At the same time, I will arrange the columns in a way that makes sense for the sentiment analysis dataset.

```{r}
#| label: sentiment-metadata-drop-columns

# Filter the data for `delivery == 'spoken'`
sotu_tbl <-
  sotu_tbl |>
  filter(delivery == "spoken") |>
  select(address_id, address_year, president, party, period, text)

# Preview
glimpse(sotu_tbl)
```

We now have a dataset with 74 addresses and then necessary metadata columns to give ample context to the sentiment analysis. The next step is to tokenize the `text` variable.

Before we do that, let's do some quick diagnostics on the dataset to make sure we understand the distribution of some of the variables. For example, we may want to know the distribution of addresses by `period`. We might also want to know if there are more addresses by a party in one period than another. Let's plot both of these distributions.

```{r}
#| label: fig-distribution-period-party
#| fig-cap: Distribution of address count and party representation by period.
#| fig-subcap:
#|  - Distribution of addresses by period
#|  - Distribution of party representation by period
#| layout-ncol: 2

# Distribution of addresses by period
sotu_tbl |>
  count(period) |>
  ggplot(aes(x = period, y = n, fill = period)) +
  geom_col() +
  labs(title = "Distribution of addresses by period")

# Distribution of party representation by period
sotu_tbl |>
  count(period, party) |>
  ggplot(aes(x = period, y = n, fill = party)) +
  geom_col(position = "dodge") +
  labs(title = "Distribution of party representation by period")
```

<!--
Note:
- The ordering of 'post' and 'pre' is alphabetical. If we want to change the order, we can make the `period` variable a factor with levels `c("pre", "post")`. At this point, we will keep the order as is as we will save this to a CSV file, and the order will not be preserved in the CSV file.
- You may also want to change the default colors in the plots. In the US the color red and blue are often associated with the Republican and Democratic parties respectively. You can change the colors using the `scale_fill_manual()` function in `ggplot2`.
-->

We can see from the plots in @fig-distribution-period-party that there are clearly more addresses pre-9/11 than post-9/11 (this make obvious sense as the dataset is from 1945 to present). We can see that the party distribution is quite similar, with a small majority of addresses from Republican presidents post-9/11.

Now, let's tokenize the `text` variable. We will use the `unnest_tokens()` function from the `tidytext` package to tokenize the `text` variable into words adding a new column `token`. I will also add a `token_id` column to keep track of the order of the tokens in the `text` variable. To do this second operation, I will group the dataset by `address_id` and then use the `row_number()` function to create the `token_id` column.

```{r}
#| label: sentiment-tokenize-text

# Tokenize the `text` variable
sotu_tbl <-
  sotu_tbl |>
  unnest_tokens(token, text) |>
  group_by(address_id) |>
  mutate(token_id = row_number()) |>
  ungroup()

# Preview
glimpse(sotu_tbl)
```

We now want to join the `bing` lexicon to the tokenized words to enable use to calculate the sentiment of each word in the exploratory analysis. The `bing` lexicon is a simple lexicon that classifies words as either positive or negative. There are other lexicons available in the `textdata` package that can be used for sentiment analysis, with varying levels of complexity. We'll take the simple approach for this analysis.

Let's load the `bing` lexicon and take a quick look at the structure of the lexicon.

```{r}
#| label: sentiment-lexicon

# Load the bing lexicon
bing_lexicon <- get_sentiments("bing")

# Preview
glimpse(bing_lexicon)
```

We see that there are almost 7k words with a sentiment classification of either positive or negative in the `bing` lexicon.

Now to join the `sotu_tbl` and the `bing_lexicon` datasets. We will join the datasets on the `token` and `word` columns respectively. The `left_join()` function from the `dplyr` package will be used to join the datasets. Left joins take the dataset 'on the left' hand side of the join and add the columns from the dataset 'on the right' hand side of the join where there is a match on the join columns.

```{r}
#| label: sentiment-join-lexicon

# Join the sotu_tbl and bing_lexicon datasets
sotu_tbl <-
  sotu_tbl |>
  left_join(bing_lexicon, by = join_by("token" == "word"))

# Preview
glimpse(sotu_tbl)
```

We now have a dataset that is suitable for sentiment analysis, at least in terms of the structure. Let's get a view of the distribution of the sentiment classes.

```{r}
#| label: tbl-sentiment-freq
#| tbl-cap: Frequency table of the `sentiment` variable

# Frequency table of the `sentiment` variable
sotu_tbl |>
  count(sentiment) |>
  kable()
```

Many, actually most, word tokens do not have a sentiment classification in the `bing` lexicon. This is not surprising because many words will not: (1) be in the lexicon and (2) have a sentiment classification by the very nature of the token (*e.g.* 'the', 'and', 'of', etc.). We will need to decide how to handle these words in the analysis. For now, we will keep them in the dataset.

Finally, let's write this dataset and create a data dictionary for the dataset. The data dictionary will be useful for future reference and for sharing the dataset with others.

```{r}
#| label: sentiment-write-data

# Write the dataset
write_csv(sotu_tbl, "../data/derived/sotu_sentiment.csv")

# Create the data dictionary
create_data_dictionary(
  sotu_tbl,
  "../data/derived/sotu_sentiment_dd.csv"
)
```

The data is written to disk and the scaffold for the data dictionary is created. The structure of the `data/` directory looks like this:

```{r}
#| label: dir-structure

# List the contents of the data directory
dir_tree("../data", recurse = 2)
```

The data dictionary with the content I added to the scaffold appears in @tbl-sentiment-dd.

```{r}
#| label: tbl-sentiment-dd
#| tbl-cap: Data dictionary for the sentiment analysis dataset
#| message: false

read_csv("../data/derived/sotu_sentiment_dd.csv") |>
  kable()
```

----

To make sure that I do not share the data or datasets as part of the GitHub repository, I will add the dataset file to the `.gitignore` file. This will ensure that the data is not shared with others when the repository is shared. Note, since our analysis is reproducible, the data can be easily recreated by running the project.

You usually will add the data files to the `.gitignore` file manually. However, I will show how you can do it programmatically. Note, you will only run this once, interactively as not to append the `.gitignore` file multiple times.

```{r}
#| label: gitignore-add-data
#| eval: false

# Add the data files to the .gitignore file
cat("../data/derived/sotu_sentiment.csv", file = "../.gitignore", append = TRUE)
```

